---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# jlme

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/jlme)](https://CRAN.R-project.org/package=jlme)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/yjunechoe/jlme/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/yjunechoe/jlme/actions/workflows/R-CMD-check.yaml)
[![test-coverage](https://github.com/yjunechoe/jlme/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/yjunechoe/jlme/actions/workflows/test-coverage.yaml)
<!-- badges: end -->

Julia (mixed-effects) regression modelling from R

## Installation

You can install the development version of jlme from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("yjunechoe/jlme")
```

## Setup

```{r example}
library(jlme)
jlme_setup()
```

`{jlme}` uses `{JuliaConnectoR}` to connect to a Julia session. See [JuliaConnectoR package documentation](https://github.com/stefan-m-lenz/JuliaConnectoR) for troubleshooting related to Julia installation and configuration.

## Using `{jlme}`

Once set up, `(g)lm()` and `(g)lmer` complements in Julia are available via `jlm()` and `jlmer()`, respectively.

### Fixed effects models

`jlm()` with `lm()`/`glm()` syntax:

```{r}
# lm(mpg ~ hp, mtcars)
jlm(mpg ~ hp, mtcars)
```

Contrasts in factor columns are preserved: 

```{r}
x <- mtcars

x$am_sum <- factor(x$am)
contrasts(x$am_sum) <- contr.sum(2)

x$cyl_helm <- factor(x$cyl)
contrasts(x$cyl_helm) <- contr.helmert(3)
colnames(contrasts(x$cyl_helm)) <- c("4vs6", "4&6vs8")

jlm(mpg ~ am_sum + cyl_helm, x)
```

### Mixed effects models

`jlmer()` with `lmer()`/`glmer()` syntax:

```{r}
data("sleepstudy", package = "lme4")
# lme4::lmer(Reaction ~ Days + (Days | Subject), sleepstudy)
jlmer(Reaction ~ Days + (Days | Subject), sleepstudy, REML = TRUE)

data("VerbAgg", package = "lme4")
# lme4::glmer(r2 ~ Anger + Gender + (1 | id), VerbAgg, family = "binomial")
jlmer(r2 ~ Anger + Gender + (1 | id), VerbAgg, family = "binomial")
```

### Inspect model objects

`{broom}`-style `tidy()` and `glance()` methods for Julia regression models:

```{r}
jmod <- jlmer(Reaction ~ Days + (Days | Subject), sleepstudy, REML = TRUE)

tidy(jmod)

glance(jmod)
```

## Advanced features with `{JuliaConnectoR}`

```{r}
library(JuliaConnectoR)
```

### Inspect model objects

```{r}
# List all properties of a MixedModel object
# - Properties are accessible via `$`
juliaCall("propertynames", jmod)

# Example 1: PCA of random effects
jmod$rePCA

# Collect as an R object (NamedTuple -> named list)
juliaGet(jmod$rePCA)
```

### Create bindings to Julia libs to access more features

Use Julia(-esque) syntax from R:

```{r}
MixedModels <- juliaImport("MixedModels")

# Check singular fit
MixedModels$issingular(jmod)

# Long-form construction of `jmod`
jmod2 <- MixedModels$fit(
  MixedModels$LinearMixedModel,
  juliaEval("@formula(Reaction ~ Days + (Days | Subject))"),
  juliaPut(sleepstudy)
)

# This time in complete Julia syntax
jmod3 <- juliaEval("
  fit(
    LinearMixedModel,
    @formula(reaction ~ days + (days | subj)),
    MixedModels.dataset(:sleepstudy)
  )
")
```

Use other `MixedModels.jl` features, like [parametric bootstrapping](https://juliastats.org/MixedModels.jl/v4/bootstrap/) for robust confidence intervals:

```{r}
Random <- juliaImport("Random")
samp <- MixedModels$parametricbootstrap(
  Random$MersenneTwister(42L), # RNG
  1000L, # Number of simulations
  jmod # Model
)
samp
```

See information about the running Julia environment (e.g., the list of loaded Julia libraries) with `jlme_status()`:

```{r}
jlme_status()
```


## Tips

### Data type conversion

Be sure to pass integers to functions that expect Integer type, (e.g., the `MixedModels.parametricbootstrap()` example above):

```{r}
# library(JuliaConnectoR)
juliaPut(1)
juliaPut(1L)
```

### Performance

In practice, most of the overhead comes from transferring the data from R to Julia. If you are looking to fit many models to the same data, you can use `jl_data()` send the data to Julia first and use that to fit models.

```{r}
data_r <- mtcars

# Extra tip: keep only columns you need
data_julia <- jl_data(data_r[, c("mpg", "am")])

jlm(mpg ~ am, data_julia)
```

If your data has custom contrasts, you can use `jl_contrasts()` to also convert that to Julia first before passing it to the model.

```{r}
data_r$am <- as.factor(data_r$am)
contrasts(data_r$am) <- contr.sum(2)

data_julia <- jl_data(data_r[, c("mpg", "am")])
contrasts_julia <- jl_contrasts(data_r)

jlm(mpg ~ am, data_julia, contrasts = contrasts_julia)
```


## Acknowledgments

- The [JuliaConnectoR](https://github.com/stefan-m-lenz/JuliaConnectoR) package for powering the R interface to Julia.

- The [Julia](https://julialang.org/) packages [GLM.jl](https://github.com/JuliaStats/GLM.jl) and [MixedModels.jl](https://github.com/JuliaStats/MixedModels.jl) for fast implementations of (mixed effects) regression models.

```{r, echo = FALSE}
stop_julia()
```
